# Survival Analysis of End-Stage Renal Disease Patients

```yaml
---
  title: "Survival Analysis of End-Stage Renal Disease Patients"
author: "Agni Charalambous"
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
toc_float: true
toc_depth: 3
theme: united
highlight: tango
number_sections: true
pdf_document:
  toc: true
toc_depth: 3
---
  ```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This project was developed as part of postgraduate coursework in the MSc in Biostatistics and Health Data Science.

The analysis investigates survival outcomes among end-stage renal disease (ESRD) patients receiving either conservative treatment or haemodialysis using classical and advanced survival analysis techniques in R.

The project includes:
  
  * Descriptive statistical analysis
* Kaplan-Meier survival estimation
* Log-rank and Wilcoxon tests
* Cox proportional hazards models
* Assessment of proportional hazards assumptions
* Time-dependent covariates
* Piecewise Cox regression
* Weibull survival modelling

# Libraries

```{r}
library(survival)
library(MASS)
```

# Data Preparation

The dataset was imported and categorical variables were converted into factors for statistical analysis.

```{r}
esrd <- read.csv("esrd.csv", sep = ";")

esrd$treatmnt <- factor(esrd$treatmnt,
                        levels = c(0, 1),
                        labels = c("Conservative", "Haemodialysis"))

esrd$sex <- factor(esrd$sex,
                   levels = c("F", "M"),
                   labels = c("Female", "Male"))

esrd$copd <- factor(esrd$copd,
                    levels = c(0, 1),
                    labels = c("No", "Yes"))

esrd$dm <- factor(esrd$dm,
                  levels = c(0, 1),
                  labels = c("No", "Yes"))

esrd$hypert <- factor(esrd$hypert,
                      levels = c(0, 1),
                      labels = c("No", "Yes"))

esrd$heart <- factor(esrd$heart,
                     levels = c(0, 1),
                     labels = c("No", "Yes"))

esrd$liver <- factor(esrd$liver,
                     levels = c(0, 1),
                     labels = c("No", "Yes"))

esrd$neoplasia <- factor(esrd$neoplasia,
                         levels = c(0, 1),
                         labels = c("No", "Yes"))

esrd$vascular <- factor(esrd$vascular,
                        levels = c(0, 1),
                        labels = c("No", "Yes"))
```

## Data Quality Checks

The dataset was screened for missing values and invalid survival times.

```{r}
str(esrd)

sapply(esrd[, c("treatmnt", "dthtime", "sex", "copd", "dm",
                "hypert", "heart", "liver", "neoplasia",
                "vascular", "dth")],
       function(x) sum(is.na(x)))

summary(esrd$dthtime)

sum(esrd$dthtime < 0, na.rm = TRUE)
```

# Descriptive Statistics

Descriptive analyses were conducted to summarize baseline characteristics and compare treatment groups.

```{r}
table(esrd$sex)
round(100 * prop.table(table(esrd$sex)), 2)

table(esrd$treatmnt)
round(100 * prop.table(table(esrd$treatmnt)), 2)

table(esrd$dth)
round(100 * prop.table(table(esrd$dth)), 2)

mean(esrd$age)
sd(esrd$age)

mean(esrd$dthtime)
sd(esrd$dthtime)
```

## Group Comparisons

Chi-square tests and t-tests were used to evaluate differences between treatment groups.

```{r}
chisq.test(table(esrd$treatmnt, esrd$sex))
chisq.test(table(esrd$treatmnt, esrd$copd))
chisq.test(table(esrd$treatmnt, esrd$dm))
chisq.test(table(esrd$treatmnt, esrd$hypert))

t.test(age ~ treatmnt, data = esrd)
t.test(dthtime ~ treatmnt, data = esrd)
```

# Risk Estimation

Cumulative incidence proportions and risk ratios were estimated for both treatment groups.

```{r}
n_con <- sum(esrd$treatmnt == "Conservative")
d_con <- sum(esrd$treatmnt == "Conservative" & esrd$dth == "1")

n_hd <- sum(esrd$treatmnt == "Haemodialysis")
d_hd <- sum(esrd$treatmnt == "Haemodialysis" & esrd$dth == "1")

p_con <- d_con / n_con
p_hd <- d_hd / n_hd

round(100 * c(conservative = p_con,
              haemodialysis = p_hd), 2)

prop.test(x = c(d_hd, d_con),
          n = c(n_hd, n_con))
```

# Mortality Rates

Mortality rates per person-year were estimated using Poisson methods.

```{r}
py_by_group <- tapply(esrd$dthtime,
                      esrd$treatmnt,
                      sum)

deaths_by_group <- tapply(esrd$dth,
                          esrd$treatmnt,
                          sum)

mr_by_group <- deaths_by_group / py_by_group

poisson.test(deaths_by_group,
             py_by_group)
```

# Kaplan-Meier Survival Analysis

Kaplan-Meier survival curves were estimated for each treatment group.

```{r}
esrd.fit <- survfit(Surv(dthtime, dth) ~ treatmnt,
                    data = esrd)

plot(esrd.fit,
     mark.time = FALSE,
     main = "Kaplan-Meier Survival Curves",
     ylab = "Survival Probability",
     xlab = "Time (years)",
     lty = 1:2,
     col = c("black", "red"))

legend("topright",
       lty = 1:2,
       col = c("black", "red"),
       legend = c("Conservative", "Haemodialysis"),
       bty = "n")
```

## Survival Curve Comparisons

Log-rank and Wilcoxon tests were used to compare survival distributions.

```{r}
survdiff(Surv(dthtime, dth) ~ treatmnt,
         data = esrd)

survdiff(Surv(dthtime, dth) ~ treatmnt,
         data = esrd,
         rho = 1)
```

# Cox Proportional Hazards Models

## Univariate Cox Model

```{r}
cox_treat <- coxph(Surv(dthtime, dth) ~ treatmnt,
                   data = esrd)

summary(cox_treat)
```

## Proportional Hazards Assumption

The proportional hazards assumption was evaluated using graphical methods and Schoenfeld residuals.

```{r}
plot(esrd.fit,
     mark.time = FALSE,
     fun = "cloglog",
     main = "Assessment of the Proportional Hazards Assumption",
     xlab = "Time (years)",
     ylab = "log(-log(Survival))",
     lty = 1:2,
     col = c("blue", "red"))

cox.zph(cox_treat)
```

# Time-Dependent Covariates

A Cox model with time-dependent treatment effects was fitted.

```{r}
esrd$trt_num <- 1 * (esrd$treatmnt == "Haemodialysis")

fit_tdc <- coxph(
  Surv(dthtime, dth) ~ trt_num + tt(trt_num),
  data = esrd,
  tt = function(x, t, ...) {x * t}
)

summary(fit_tdc)
```

# Piecewise Cox Regression

Piecewise Cox regression was used to allow treatment effects to vary across time intervals.

```{r}
cuts <- c(1, 3)

esrd_split <- survSplit(
  Surv(dthtime, dth) ~ .,
  data = esrd,
  cut = cuts,
  episode = "interval",
  zero = -1
)

esrd_split$interval <- factor(
  esrd_split$interval,
  labels = c("0-1", "1-3", ">=3")
)

fit_piecewise <- coxph(
  Surv(tstart, dthtime, dth) ~ treatmnt * interval,
  data = esrd_split
)

summary(fit_piecewise)
```

# Multivariable Cox Regression

A multivariable Cox proportional hazards model was fitted including demographic and clinical covariates.

```{r}
cox_mult <- coxph(
  Surv(dthtime, dth) ~
    treatmnt + age + sex + copd + dm + hypert +
    heart + liver + neoplasia + vascular,
  data = esrd
)

summary(cox_mult)
```

# Weibull Survival Model

Parametric survival analysis was performed using Weibull regression models.

```{r}
esrd$dthtime_fixed <- esrd$dthtime

esrd$dthtime_fixed[esrd$dthtime_fixed <= 0] <- 0.001

fit_weibull <- survreg(
  Surv(dthtime_fixed, dth) ~ 1,
  data = esrd,
  dist = "weibull"
)

summary(fit_weibull)
```

# Conclusion

This project demonstrates the application of classical and advanced survival analysis techniques in the evaluation of treatment outcomes among end-stage renal disease patients.

The analysis included non-parametric, semi-parametric, and parametric survival modelling approaches using R. The project provided practical experience in statistical modelling, survival analysis methodology, model diagnostics, and interpretation of healthcare-related data.
